\bf \Large A Model Evaluation Study for Treatment Planning of Laser Induced Thermal Therapy

### 2015-04-14 gpuPlanningAug13.tex
Revised fig 3 and caption and Reviewer Response

Previous real-time implementations of the kernel for
predicting the laser induced bioheat transfer were made feasible via
implementation on a massively parallel supercomputers~\cite{fuentesetal09,
Odenetal07,Dilleretal07,Fengetal07,Fuentesetal08}. 
Despite the increasing presence of large computational clusters within
large academic hospitals, reliance on such an approach makes clinical
translation of this research extremely difficult. This technology must
be practically available to any clinic using an MR-guided thermal therapy
system whereas integration with an institution's cluster and intranet creates
significant overhead and potential complications.
While modern single-chip multi-core CPU's have a peak performance of
$\mathcal{O}(10-100)$ GFLOPs and are sufficient 
for inexpensive algorithms such as the linear superposition of analytic functions presented
in Section~\ref{AnalyticBioheatModel}, 
current GPUs~\cite{Farber2011a,Muller2013} have a peak
performance of $\mathcal{O}(1-2)$ TFLOPs (single precision) and the
portability of these devices are ideal for developing more advanced 
numerical discretizations of the underlying partial differential equations
for guiding the therapy. However, a significant
time investment is needed to express the class of memory bound numerical
methods used in simulation within an algorithmic framework that can
simultaneously exploit these floating point capabilities and minimize the
global memory access latency of these devices.
% 28 = 1030/ (148/4) single precision.
% max floating point operations per second is 1030 GFlop
% data transfer of single precision numbers to the cores = bandwidth x data type size = 148  x (4bytes / float)
% >>> 1030./ (148./4.)
% 27.83783783783784
For example, the $\mathcal{O} (150)$~GB/s memory bandwidth of the Tesla M2070
used in the manuscript requires $\approx$ 30 floating point operations per
single precision global memory access to fully utilize the floating point capabilities
of the GPU. 
The spectral element method presented in Section~\ref{BioheatComputerModel}
achieves efficient global memory access through a matrix-free
implementation of the inherent linear system of equations.
As opposed to explicitly storing and reading matrix entries
from global memory, the matrix-free method recalculates the local matrix entries
needed within the linear system solve.
This memory access design pattern has demonstrated a 4-10x speedup  over
the matrix explicit methods~\cite{Muller2013,Knepley2014,Medina2014}.

@@

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure*}[p!]
\centering
\iflatextortf 
%do nothing
\else
\begin{tabular}{ccc}
\scalebox{0.26}{\includegraphics*{\picdir/OverviewApplicator}}  
&
\scalebox{0.18}{\includegraphics*{\picdir/FEMesh}}  
&
\scalebox{0.24}{\includegraphics*{\picdir/powerProfile}}  
\\
(a) & (b) & (c)  \\
\end{tabular}
\fi
\caption{
(a) The Visualase\textsuperscript{\textregistered} applicator modeled in
this application and a diagram of the photon emitting diffusing
tip and the cooling fluid  are shown. (b) A finite element mesh conforms to the applicator
and is used as the template for the calculations. (c) A representative
time-temperature history profile of the thermometry data at two points within the brain tissue,
 $\sim$1 mm from the applicator, is shown. The corresponding power history is also shown. 
} \label{LaserApplicatorOverview}
\end{figure*}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

MR temperature imaging was performed on a 1.5 T MRI
(GE Healthcare Technologies, Waukesha, WI)
using a 2D gradient echo sequence~\cite{staffordetal04} 
(FA = \SI{30}{\degree}, FOV = $24 \times 24$ cm\textsuperscript{2}, matrix size = $256 \times 128$, 
TR/TE=37.5/20 ms, receive-only head coil,  
5 sec. per update).
%No acceleration factor was used.
The imaging plane was chosen perpendicular to the axial direction of the applicator 
and allowed monitoring of critical structure regions.
The Visualase\textsuperscript{\textregistered} workstation communicates with the MR scanner to obtain
raw DICOM imaging data during the procedure.
The temperature dependent water proton resonance frequency shift is measured
by calculating the complex phase-difference observed during heating.
The water proton resonance frequency chemically shifts to lower
frequencies with higher temperatures (caused by rupture,
stretching, bending of hydrogen bonds)~\cite{ishihara1995paf}.
The total
temperature change, $\Delta u$, is proportional to the measured
phase change, $\delta \phi$.
\[
\Delta u = \frac{ \delta \phi } { 2 \pi \alpha \cdot \gamma B_0 \cdot \text{TE}}
\]
%$\alpha = -0.0097 \left[\frac{ppm}{^oC}\right]$ 
Here $\alpha$ is the temperature sensitivity coefficient, $\gamma$ is the
gyromagnetic ratio of water, $B_0$ is the static magnetic field strength, and
TE is the echo time.  
A baseline body temperature of \textit{u}\textsubscript{0} = \SI{37.0}{\degreeCelsius} is assumed to obtain
absolute temperature.
An Arrhenius rate process model~\cite{pearce1995rate}
was used to evaluate the thermal dose resulting from the
time-temperature history of the laser exposure.
\begin{equation}\label{ArrheniusDamage}
\text{$\Omega$}(t) = 
      \int_0^t \! Ae^{\frac{-E_A}{Ru(\tau)}} \, \mathrm{d} \tau
\end{equation}
In this Arrhenius thermal dose model, the frequency factor, $A$,
and the activation energy, $E_A$, are 
experimentally determined kinetic parameters. The values for $A$ and
$E_A$ were 3.1E98 and 6.28E5,
respectively, and have been used in previous
studies~\cite{carpentieretal08,schwarzmaier1998treatment,mcnichols2004mr}. 
$R$ is the universal gas constant.  The thermal dose was assumed to
be lethal at doses $\omega$  $\geq 1$ as seen in previous
reports~\cite{carpentieretal08,mcnichols2004mr}.

@@

\begin{figure*}[p!]
\begin{tabular}{ccccc} 
\centering
\iflatextortf 
%do nothing
\else
%\begin{tabular}{cc}
%\scalebox{0.42}{\includegraphics*{\picdir/datasummarybestfit}}  
%&
\scalebox{0.55}{\includegraphics*{\picdir/hist_pics/FEM_Hist.png}} &  \scalebox{0.55}{\includegraphics*{\picdir/hist_pics/SteadyStateHist22.png}}  
\\
(a) & (b) \\
\fi
\end{tabular}
\caption{ 
Presented here are histograms of calibration analysis from the transient FEM model (a) and steady state model (b),
shown left and right; respectively. The optical parameters, $\mu_\textit{eff}$,
recovered from each thermometry dataset considered is shown.
For each calibration, the bound constrained optimization was
restricted to a range obtained from literature, $\mu_\textit{eff} \in [0.8,400]$ m$^{-1}$. Leave-one-out cross validation was performed
using these \numberofpatientsLOOCV $\mu_\textit{eff}$ values.
The nominal value in brain tissue obtained from~\cite{Welch95} is 
$\mu_\textit{eff}$ = 180 m$^{-1}$.
} \label{QuantitativeComparison}
\end{figure*}

>>>>

Previous real-time implementations of the kernel for
predicting the laser induced bioheat transfer were made feasible via
implementation on a massively parallel supercomputers~\cite{fuentesetal09,
Odenetal07,Dilleretal07,Fengetal07,Fuentesetal08}. 
Despite the increasing presence of large computational clusters within
large academic hospitals, reliance on such an approach makes clinical
translation of this research extremely difficult. This technology must
be practically available to any clinic using an MR-guided thermal therapy
system whereas integration with an institution's cluster and intranet creates
significant overhead and potential complications.
While modern single-chip multi-core CPU's have a peak performance of
$\mathcal{O}(10-100)$ GFLOPs and are sufficient 
for inexpensive algorithms such as the linear superposition of analytic functions presented
in Section~\ref{AnalyticBioheatModel}, 
current GPUs~\cite{Farber2011a,Muller2013} have a peak
performance of $\mathcal{O}(1-2)$ TFLOPs (single precision) and the
portability of these devices are ideal for developing more advanced 
numerical discretizations of the underlying partial differential equations
for guiding the therapy. However, a significant
time investment is needed to express the class of memory bound numerical
methods used in simulation within an algorithmic framework that can
simultaneously exploit these floating point capabilities and minimize the
global memory access latency of these devices.
% 28 = 1030/ (148/4) single precision.
% max floating point operations per second is 1030 GFlop
% data transfer of single precision numbers to the cores = bandwidth x data type size = 148  x (4bytes / float)
% >>> 1030./ (148./4.)
% 27.83783783783784
For example, the $\mathcal{O} (150)$~GB/s memory bandwidth of the Tesla M2070
used in the manuscript requires $\approx$ 30 floating point operations per
single precision global memory access to fully utilize the floating point capabilities
of the GPU. 
The spectral element method presented in Section~\ref{BioheatComputerModel}
achieves efficient global memory access through a matrix-free
implementation of the inherent linear system of equations.
As opposed to explicitly storing and reading matrix entries
from global memory, the matrix-free method recalculates the local matrix entries
needed within the linear system solve.
This memory access design pattern has demonstrated a 4-10x speedup  over
the matrix explicit methods~\cite{Muller2013,Knepley_2_2013,Medina2014}.

@@

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure*}[p!]
\centering
\iflatextortf 
%do nothing
\else
\begin{tabular}{ccc}
\scalebox{0.4}{\includegraphics*{\picdir/OverviewApplicator}} \\ 
(a)\\
\scalebox{0.18}{\includegraphics*{\picdir/FEMesh}}  
&
\scalebox{0.24}{\includegraphics*{\picdir/powerProfile}}  
\\
(b) & (c)  \\
\end{tabular}
\fi
\caption{
(a) The Visualase\textsuperscript{\textregistered} applicator modeled in
this application and a diagram of the photon emitting diffusing
tip and the cooling fluid  are shown. (b) A finite element mesh conforms to the applicator
and is used as the template for the calculations. (c) A representative
time-temperature history profile of the thermometry data at two points within the brain tissue,
 $\sim$1 mm from the applicator, is shown. The corresponding power history is also shown. 
} \label{LaserApplicatorOverview}
\end{figure*}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

MR temperature imaging was performed on a 1.5 T MRI
(GE Healthcare Technologies, Waukesha, WI)
using a 2D gradient echo sequence~\cite{staffordetal04} 
(FA = \SI{30}{\degree}, FOV = $24 \times 24$ cm\textsuperscript{2}, matrix size = $256 \times 128$, 
TR/TE=37.5/20 ms, receive-only head coil,  
5 sec. per update).
%No acceleration factor was used.
The imaging plane was chosen perpendicular to the axial direction of the applicator 
and allowed monitoring of critical structure regions.
The Visualase\textsuperscript{\textregistered} workstation communicates with the MR scanner to obtain
raw DICOM imaging data during the procedure.
The temperature dependent water proton resonance frequency shift is measured
by calculating the complex phase-difference observed during heating.
The water proton resonance frequency chemically shifts to lower
frequencies with higher temperatures (caused by rupture,
stretching, bending of hydrogen bonds)~\cite{ishihara1995paf}.
The total
temperature change, $\Delta u$, is proportional to the measured
phase change, $\delta \phi$.
\[
\Delta u = \frac{ \delta \phi } { 2 \pi \alpha \cdot \gamma B_0 \cdot \text{TE}}
\]
%$\alpha = -0.0097 \left[\frac{ppm}{^oC}\right]$ 
Here $\alpha$ is the temperature sensitivity coefficient, $\gamma$ is the
gyromagnetic ratio of water, $B_0$ is the static magnetic field strength, and
TE is the echo time.  
A baseline body temperature of \textit{u}\textsubscript{0} = \SI{37.0}{\degreeCelsius} is assumed to obtain
absolute temperature.
An Arrhenius rate process model~\cite{pearce1995rate}
was used to evaluate the thermal dose resulting from the
time-temperature history of the laser exposure.
\begin{equation}\label{ArrheniusDamage}
\text{$\Omega$}(t) = 
      \int_0^t \! Ae^{\frac{-E_A}{Ru(\tau)}} \, \mathrm{d} \tau
\end{equation}
In this Arrhenius thermal dose model, the frequency factor, $A$,
and the activation energy, $E_A$, are 
experimentally determined kinetic parameters. The values for $A$ and
$E_A$ were 3.1E98 {\color{red}s\textsuperscript{-1}} and 6.28E5 {\color{red}(J/mol)},
respectively, and have been used in previous
studies~\cite{carpentieretal08,schwarzmaier1998treatment,mcnichols2004mr}. 
$R$ is the universal gas constant.  The thermal dose was assumed to
be lethal at doses $\omega$  $\geq 1$ as seen in previous
reports~\cite{carpentieretal08,mcnichols2004mr}.

@@

\begin{figure*}[p!]
\begin{tabular}{ccccc} 
\centering
\iflatextortf 
%do nothing
\else
%\begin{tabular}{cc}
%\scalebox{0.42}{\includegraphics*{\picdir/datasummarybestfit}}  
%&
\scalebox{0.55}{\includegraphics*{\picdir/hist_pics/FEM_Hist.png}} &  \scalebox{0.55}{\includegraphics*{\picdir/hist_pics/SteadyStateHist22.png}}  
\\
(a) & (b) \\
\fi
\end{tabular}
\caption{ 
Presented here are histograms of calibration analysis from the transient FEM model (a) and steady state model (b),
shown left and right; respectively. {\color{red}Both histograms have a bin width of \SI{13.0}{\per\metre}.}
The optical parameters, $\mu_\textit{eff}$,
recovered from each thermometry dataset considered is shown.
For each calibration, the bound constrained optimization was
restricted to a range obtained from literature, $\mu_\textit{eff} \in [0.8,400]$ m$^{-1}$. Leave-one-out cross validation was performed
using these \numberofpatientsLOOCV $\mu_\textit{eff}$ values.
The nominal value in brain tissue obtained from~\cite{Welch95} is 
$\mu_\textit{eff}$ = 180 m$^{-1}$.
} \label{QuantitativeComparison}
\end{figure*}
